Reproducible Research Peer Assignment #1
==========================
  
# Data Description
  
The data can be downloaded from the course website ([link][1]). This data was the number 
of steps taken in 5 minute intervals each day, taken from an anonymous individual using a 
personal activity monitoring device (like a [Fitbit][2], [Nike Fuelband][3], or [Jawbone Up][4]). 
This is two months worth of data from this individual, taken during the months of October and November.  

It contains three columns of data:
- **steps**: Number of steps taken in a five minute interval (missing values are "NA")
- **date**: The date on which the measurement was taken in YYYY-MM-DD format
- **interval**: The start of the 5-minute interval in which the measurement was taken (in military time)  

This document describes an analysis performed on this data.  
  
# Step 1: Load and preprocess the data
  
## Load the data

To handle situations where the data is not in the same directory as the working one, I usually
prompt the user to give the program the directory in which the data resides:

```{r promptDir}
activity.table <- read.csv("activity.csv")
```
  
## Number of NA values and where are they?

A quick check shows that there are `r sum(is.na(activity.table$steps))` NA values for "steps" 
taken and that there are 288 measurements for each day of recording (which is equal to the 
number of 5 minute intervals in the day: 24 hrs * 60 minutes / 5 = 288).  

```{r summaryStats}
summary(activity.table$steps)
summary(activity.table$date)
```
  
## How will we transform the data to do the analysis?

To do the last section, I needed to add a column for weekday. I do this by utilizing the POSIXlt class
to convert the date to a date object, and then extracting the weekday from that object. In
addition, in order to compare weekdays to weekends, I created a variable that coded that as well. 
  
```{r addWeekdays}
activity.table$weekday <- weekdays(as.POSIXlt(activity.table$date))
activity.table$dayType <- as.factor(activity.table[,"weekday"] %in% c("Sunday", "Saturday"))
levels(activity.table$dayType) <- c("weekday","weekend")
```
  
To do the analysis of total steps per day and average over each interval, I need to create
respective tables that aggregate along each day and along each interval. In addition, to 
fill in the NA values later on, I need a table that aggregates along both weekday and
interval.  

```{r totalStepsPerDay}
totalStepsPerDay <- aggregate(activity.table$steps, by=list(activity.table$date), sum)
avgStepsPerInterval <- aggregate(activity.table$steps, by=list(activity.table$interval), mean, na.rm=T)
avgStepsPerWeekday <- aggregate(activity.table$steps, by=list(activity.table$weekday, activity.table$interval), mean, na.rm=T)
```

```{r renaming, echo=F, results="hide"}
names(totalStepsPerDay) <- c("date","totalSteps")
names(avgStepsPerInterval) <- c("interval","avgSteps")
names(avgStepsPerWeekday) <- c("weekday","interval","avgSteps")
```
  
## Summary of transformations
  
This is a sample of what the tables look like after this transformation:
```{r summaryTables, echo=F}
head(activity.table)
head(totalStepsPerDay)
head(avgStepsPerInterval)
head(avgStepsPerWeekday)
```
  
# Step 2: Mean Total Number of Steps Per Day
  
## Histogram of the Total Steps Taken Per Day
  
```{r totalStepsPlot, echo=F}
library(ggplot2)
qplot(as.POSIXct(totalStepsPerDay$date),totalStepsPerDay$totalSteps,geom="histogram",stat="identity",xlab="Date",ylab="Total Steps", na.rm=T)
```

## Mean and Median Total Steps Taken Per Day

```{r meanAndMedian}
mean(totalStepsPerDay$totalSteps, na.rm=T)
median(totalStepsPerDay$totalSteps, na.rm=T)
```

# Step 3: What is the average daily activity pattern?

### Time Series Plot of Average Steps Taken Over Each 5-minute Interval Across All Days
Here is a time series plot of the average steps taken during each 5-minute interval across the day:
```{r avgStepsAcrossDay, echo=F}
plot(avgStepsPerInterval$interval, avgStepsPerInterval$avgSteps, type="l", xlab="5-minute Interval (Military Time)", ylab="Avg Steps", xaxt="n")
axis(1, at=seq(0,2400,100),labels=seq(0,2400,100))
```

### The 5-minute interval with the most average number of steps?

The 5-minute interval with the maximum number of steps, as seen in the graph, was 
`r avgStepsPerInterval[max(avgStepsPerInterval$avgSteps),"interval"]`.

# Step 4: Imputing Missing Values
  
As mentioned above, I decided to impute missing values based on the average steps for an interval. Here is how I would do that:
  
```{r imputeMissingValues}
updated.table <- activity.table
updated.table[is.na(updated.table$steps),"steps"] <- apply(updated.table[is.na(updated.table$steps),], 1, function(x) avgStepsPerInterval[which(avgStepsPerInterval$interval==as.integer(x[3])),"avgSteps"])
```

# Step 5: Weekday vs Weekend


[1]: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip "Data"
[2]: http://www.fitbit.com/ "Fitbit"
[3]: http://www.nike.com/us/en_us/c/nikeplus-fuelband "Nike"
[4]: https://jawbone.com/up "Jawbone"